Equivalence Testing and the Second Generation P-Value

### 2018-07-28 sgpv_vs_TOST.Rmd
Update discussing 4 situations when p-values from TOST and SGPV are unrelated

---
title: "Equivalence Testing and the Second Generation P-Value"
author: "Daniël Lakens & Marie Delacre"
date: "26 juli 2018"
output: github_document
---

@@

  invisible(capture.output(res <- TOSTone.raw(m = m, 
                                              mu = mu,
                                              sd = sd, 
                                              n = n, 
                                              low_eqbound = low_eqbound, 
                                              high_eqbound = high_eqbound, 
                                              alpha = alpha,
                                              plot = FALSE
  )))
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```

```{r sgpv_tost, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 1001),
     yaxt = "n",
     xaxt = "n",
     ylab = "",
     xlab = "Mean")
axis(1, at = seq(0,1000,100), labels = seq(140,150,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

@@

>>>>

---
title: "Equivalence Testing and the Second Generation P-Value"
author: "Daniël Lakens & Marie Delacre"
date: "28 juli 2018"
output: github_document
---

@@

  invisible(capture.output(res <- TOSTone.raw(m = m, 
                                              mu = mu,
                                              sd = sd, 
                                              n = n, 
                                              low_eqbound = low_eqbound, 
                                              high_eqbound = high_eqbound, 
                                              alpha = alpha,
                                              plot = FALSE
  )))
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r sgpv_tost, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 1001),
     yaxt = "n",
     xaxt = "n",
     ylab = "",
     xlab = "Mean")
axis(1, at = seq(0,1000,100), labels = seq(140,150,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

@@

##When are the SGPV and Equivalence Test Unrelated?

There are 4 situations where *p*-values from TOST and SGPV are unrelated. The first two situations were discussed earlier, and can be seen in Figure  1. When the SGPV is either 0 or 1 *p*-values from the equivalence test fall between 0.975 and 1 or between 0 and 0.025. Because p-values approach 0 or 1, but are never exactly 0 or 1, while the SGPV is exactly 0 or 1, the two statistics are completely unrelated. The easiest way to see this is by plotting the SGPV against the *p*-value from the TOST procedure. The situations where the SPGV and *p*-values from the TOST procedure are unrelated are indicated by the parts of the curve where there are vertical lines at SGPV of 0 and 1.
```{r, include=FALSE}
step = 0.01

  invisible(capture.output(res <- TOSTone.raw(m = m, 
                                              mu = mu,
                                              sd = sd, 
                                              n = n, 
                                              low_eqbound = low_eqbound, 
                                              high_eqbound = high_eqbound, 
                                              alpha = alpha,
                                              plot = FALSE
  )))
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 5*: The relationship between *p*values from the TOST procedure and the SGPV for the same scenario as in Figure 1.

A third situation in which the SGPV deviates strongly from the TOST *p*-value is whenever the CI is more than twice as wide as the equivalence range, and the CI overlaps with the upper *and* lower equivalence bound. In this situation the normal calculation of the proportion of overlap is skipped, and the SGPV is set to 0.5 instead. Blume et al. (2018) call this the 'small sample correction factor'. However, it is not a correction in the typical sense of the word, since the SGPV is not adjusted to any 'correct' value. When the normal calculation would be 'misleading' (i.e., the SGPV would be small, which normally would suggest support for the alternative hypothesis, when all values in the equivalence range are also supported), the SGPV is set to 0.5 which according to Blume and colleagues signal the SGPV is 'uninformative'. Note that the CI can be twice as wide as the equivalence range whenever the sample size is small (and the confidence interval width is large) *or* when then equivalence range is narrow. It is therefore not so much a 'small sample correction' as it is an exception to the typical calculation of the SGPV whenever the ratio of the confidence interval width to the equivalence range exceeds 2:1 and the CI overlaps with the upper and lower bounds.

We can examine this situation be calculating the SGPV and performing the TOST for a situation where sample sizes are small and the equivalence range is narrow, such that the CI is more than twice as large as the equivalence range. 
```{r, include=FALSE}
step = 0.01

p_tost_list <- numeric(length(seq(-3, 3, step)))
sgpv_list <- numeric(length(seq(-3, 3, step)))
p_list <- numeric(length(seq(-3, 3, step)))
t_list <- numeric(length(seq(-3, 3, step)))

for(i in seq(-3, 3, step)){
  count <- count + 1
  m <- i
  mu <- 0
  sd <- 2
  n <- 10
  low_eqbound = -0.4 
  high_eqbound = 0.4
  alpha = 0.05

  invisible(capture.output(res <- TOSTone.raw(m = m, 
                                              mu = mu,
                                              sd = sd, 
                                              n = n, 
                                              low_eqbound = low_eqbound, 
                                              high_eqbound = high_eqbound, 
                                              alpha = alpha,
                                              plot = FALSE
  )))
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 600),
     yaxt = "n",
     xaxt = "n",
     ylab = "",
     xlab = "Mean")
axis(1, at = seq(0,600,100), labels = seq(-3,3,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

lines(sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
abline(h = 0.5, 
       lty = 6,
       col = "lightgrey")

```
*Figure 6*: Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of true population means (x-axis). Because the sample size is small (n = 10) and the CI is more than twice as wide as the equivalence range (set to -0.4 to 0.4), the SGPV is set to 0.5 (horizontal lightgrey line) across a range of observed means.

We can again plot the two statistics against each other so see where they are unrelated (indicated by staight lines in the curve). We see the SGPV is 0.5 for a range of observed means where the p-value from the equivalence test still varies. It should be noted that in these calculations the *p*-values for the TOST procedure are *never* smaller than 0.05 (i.e., they do not get below 0.05 on the y-axis). In other words, we can't conclude equivalence based on any of the observed means. How is this possible? Remember that the TOST procedure consists of two one-sided tests against the upper and lower equivalence bound. The TOST *p*-value is smaller than 0.05 if the 90% CI falls completely between the upper and lower equivalence bounds. However, we are examining a scenario where the 90% CI is so wide that it never falls completely within the two equivalence bounds. As Lakens (2017) notes: “in small samples (where CIs are wide), a study might have no statistical power (i.e., the CI will always be so wide that it is necessarily wider than the equivalence bounds).” None of the *p*-values based on the TOST procedure are below 0.05, and thus, in the long run we have 0% power.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 7*: The relationship between *p*values from the TOST procedure and the SGPV for the same scenario as in Figure 6.

There is one last situation where the *p*-value from the TOST procedure and the SGPV are unrelated. This is when the CI overlaps with the upper and lower equivalence bound, but the CI is *not* twice as wide as the equivalence range. This fourth category exists because of the decision by Blume and collagues to set the SGPV to 0.5 whenever the CI is twice as wide as the equivalence range and the CI overlaps with both equivalence bounds. This means that there are situations where the CI interval overlaps with both equivalence bounds, while the CI is less than twice as large as the equivalence bound. For example, in the example below, we see an example where the CI is only 1.79 times as wide as the equivalence bounds, but the CI overlaps with the lower and upper equivalence bounds.
```{r, include=FALSE}
m <- 0.2
n <- 10
sd <- 1
d.lo <- -0.4
d.hi <- 0.4

tost_res1 <- TOSTone.raw(m = m, 
                         mu = 0,
                         sd = sd,
                         n = n,
                         low_eqbound = d.lo,
                         high_eqbound = d.hi,
                         alpha = 0.05,
                         plot = F)

#Ratio CI to equivalence bounds
(tost_res1$UL_CI_TTEST - tost_res1$LL_CI_TTEST) / (d.hi - d.lo)

```
```{r, echo=FALSE, dpi=600, fig.width=5, fig.height=4}

plot(NA, 
     ylim = c(0, 1), 
     xlim = c(-2, 2),
     yaxt = "n",
     ylab = "SGPV or TOST p-value",
     xlab = "Mean Difference")
axis(1, at = c(-3,-2,-1,0,1,2,3,4,5), las = 1)
abline(v = tost_res1$high_eqbound, 
       lty = 2)
abline(v = tost_res1$low_eqbound, 
       lty = 2)
abline(v = 0, 
       lty = 2, 
       col = "grey")

points(x = tost_res1$diff, 
       y = 0.5, 
       pch = 15, 
       cex = 2)
segments(tost_res1$LL_CI_TTEST, 
         0.5, 
         tost_res1$UL_CI_TTEST, 
         0.5, 
         lwd = 3)
```
*Figure 8*: Example of a 95% CI that overlaps with the lower and upper equivalence bound (indicated by the vertical dotted lines).

If the observed mean would be somewhat closer to 0, or further away from 0, the SGPV would remain constant (the CI width does not change, it ccompletely overlaps with the equivalence range) while the *p*-value for the TOST procedure varies. We can see this in Figure 9 below. The SGPV is not set to 0.5, but is slightly higher than 0.5 across a range of means. How high the SGPV will be when the CI overlaps with the lower and upper equivalence bounds, but the CI is not twice as large as the equivalence range, depends on the width of the CI and the equivalence range. 
```{r, include=FALSE}
step = 0.01

p_tost_list <- numeric(length(seq(-3, 3, step)))
sgpv_list <- numeric(length(seq(-3, 3, step)))
p_list <- numeric(length(seq(-3, 3, step)))
t_list <- numeric(length(seq(-3, 3, step)))

for(i in seq(-3, 3, step)){
  count <- count + 1
  m <- i
  mu <- 0
  sd <- 1
  n <- 10
  low_eqbound = -0.4 
  high_eqbound = 0.4
  alpha = 0.05

  invisible(capture.output(res <- TOSTone.raw(m = m, 
                                              mu = mu,
                                              sd = sd, 
                                              n = n, 
                                              low_eqbound = low_eqbound, 
                                              high_eqbound = high_eqbound, 
                                              alpha = alpha,
                                              plot = FALSE
  )))
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 600),
     yaxt = "n",
     xaxt = "n",
     ylab = "",
     xlab = "Mean")
axis(1, at = seq(0,600,100), labels = seq(-3,3,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

lines(sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
abline(h = 0.5, 
       lty = 6,
       col = "lightgrey")
```
*Figure 9*: Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of true population means (x-axis). The sample size is small (n = 10), but because the sd is half as big as in Figure 7 (1 instead of 2) the CI is less than twice as wide as the equivalence range (set to -0.4 to 0.4). The SGPV is not set to 0.5 (horizontal lightgrey line) but reaches a maximum slightly above 0.5 across a range of observed means.

If we once more plot the two statistics against each other so see where they are unrelated (indicated by staight lines in the curve). We see the SGPV is 0.56 for a range of observed means where the *p*-value from the equivalence test still varies.
```{r, echo=FALSE, dpi=600, fig.width=6, fig.height=5}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```
*Figure 10*: The relationship between *p*values from the TOST procedure and the SGPV for the same scenario as in Figure 6.

To conclude this section, there are four situations where the *p*-value from the TOST procedure is unrelated to the SGPV, In all these situation the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely descriptive statistic, the SGPV is more limited than the *p*value from the TOST procedure. The proportion of overlap can be the same value when the observed mean is 0 or when the observed mean falls just inside the equivalence bound, and additional information (e.g., the 95% CI) is required to differentiate these situations. One way to mitigate this limitation of the SGPV would be to set the SGPV to 0.5 whenever the CI overlaps with both the upper and lower equivalence bound (irrespective of the width of the CI).
